Introduction: a pipeline to analyse time-course gene expression data

### 2019-04-23 scripts/00_introduction.Rmd
Merging Elizabeth's introduction to the Rmd manuscript

@@

This workflow relies on several standard packages for analysing gene
expression data, some specific for time-course data, others broadly used by
the community. We also provide alongside this workflow a package that nicely
wraps all calls to dependencies with a common API by defining a model object
holding all necessary metadata to perform all steps of the pipeline.

>>>>

Gene expression studies provide simultaneous quantification of the level of
mRNA from all genes in a sample. High-throughput studies of gene expression
have a long history, starting with microarray technologies in the 1990s
through to single-cell technologies. While many expression studies are
designed to compare the gene expression in distinct groups, there is also a
long history of time-course expression studies, where the the gene expression
is compared across time by measuring mRNA levels from different samples across
time.\footnote{Because the collection of the mRNA is often destructive,
samples at different time points are generally from different biological
samples; longitudinal studies, for example tracking the same subject over
time, are certainly possible in certain settings, but not directly considered
here.} Such time course studies can vary from measuring a few distinct time
points, to sampling ten to twenty time points. Many longer time series are
particularly interested in investigating development over time. More recently,
single-cell studies track single cells through their development, and a single
cell is measured at a particular moment in its developmental progression -- a
value that is not know but estimated from the data as its "pseudo-time."

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being
\texttt{edge}\todo[Other methods to mention: impulse models, what else?
Citations!]. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show *some* change.
For example, in the \xxx system, over \xxx\% of genes are shown to be changing
over time. The task in these settings is often not to detect changes in genes,
but to categorize into biologically interpretable patterns the vast number of
changes discovered.

We present here a workflow for such an analysis that consists of \xxx main
parts (Figure \ref{fig:workflow}):

- Identification of genes that are differentially expressed
- Clustering of genes into distinct temporal patterns
- Biological interpretation of the clusters

**FIXME EP This is probably not best enumeration of steps, but need
something...**

This workflow represents an integration of both novel implementations of
previously established methods and new methodologies for the settings of
developmental time series. We provide the various steps of the workflow as
functions in a R package, \citettt{\xxx}, along with command-line tools for
running the entire workflow in a single command.

---
# Old intro

@@

This workflow relies on several standard packages for analysing gene
expression data, some specific for time-course data, others broadly used by
the community. We also provide alongside this workflow a package that nicely
wraps all calls to dependencies with a common API by defining a model object
holding all necessary metadata to perform all steps of the pipeline.
---
